Labor Market Income and Substitution Effects Explain Modern Natural Selection

### 2020-12-08 why-negative-selection.Rmd
Work on text. Add summary of inequality effects.

---
title: "Economic Theory of Fertility Explains Modern Natural Selection"
author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. Polygenic scores which predict lower earnings and education are selected for. Effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Overall effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Times
---

@@

* Limit max no. of children in regressions (use an "in_children" bool)
  - not necessary. Very few have abnormally many.
* Consider doing separate regressions for "any children" vs. "n children | 1 or more"
  - heckit too slow and gives crazy answers, just use plm/ols
  - some variables do look different
* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1.
* Reweight other regressions (e.g. the omnibus regression in the appendix)?
* Another point is that negative selection only among poor will increase
    genetic inequality.
  - one way to look at this: regress PGS on e.g. income or Quin71
  - then weight by n_children and see the change
  - needs to be comparable in some sense though, across generations
  - e.g. quintile of where born, unweighted; vs quintile of where now,
    weighted by children
  - perhaps simpler is just to run regressions with/without weights
    by n_children. This gives you the "effect" of natural selection
* If you take "year of last job started" and see if it is after/before
  the first child, you can figure out whether mothers keep working.
  - better still, look at all jobs started and see if any are within e.g. 5 years of
  first child's birth
  - this could then be another way to get at income vs substitution effects. You'd
  predict negative selection effects among mothers who worked. Of course the
  choice to work is endogenous.
* double check which scores are reverse coded!

@@

As yet we know little about the *mechanisms* behind natural selection in
contemporary human populations. This study uses data from UK Biobank to learn
more. We test for natural selection on 33 different polygenic scores. We extend
the analysis over two generations, using data on respondents' number of siblings
as well as their number of children. This is interesting because consistent
natural selection over multiple generations could lead to substantive effects in
the long run.

Most importantly, we examine reproductive rates in different
subgroups of the population, in order to uncover patterns that can help
illuminate the mechanisms behind modern natural selection. We find that 
across the board, polygenic scores have stronger relationships with fertility 
in specific subgroups of the population. Furthermore, in some subgroups, 
the direction of relationships is actually reversed.

@@

We use data from UK Biobank. Our sample consists of white British (as defined
by genetic estimated ancestry). For regressions on number of children we 
limit the sample to people over 45, since most fertility is completed by this age.

We compute 33 polygenic scores, normalized
to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order
of 5% of a standard deviation.

@@

These changes could reflect either natural selection within the UK
population, or ascertainment bias within the sample. Respondents are
higher income and better educated than the UK population, and they may
also differ on other unobserved characteristics [@10.1093/aje/kwx246].
Since richer and educated people also live longer, this bias might also
increase with age.

We next regress respondents'polygenic scores on their number of children
($y_i$):

@@

The "selection effect", $\beta$, measures the strength of natural
selection within the sample. In fact, since polygenic scores are
normalized to mean 0, $\beta$ is the expected polygenic score among
children of the sample [@beauchamp2016genetic]. To learn more about the
underlying mechanisms, we split the sample, starting with basic
demographic variables including education, income and sex. These are all
potential sources of ascertainment bias: as well as the ascertainment
for income and education, mentioned above, the sample sex ratio skews
`r female_ratio * 100`% female.

@@

```{r calc-income-edu-controlled}
drake::loadd(res_income_controlled, res_edu_controlled)
n_regs_ic <- nrow(res_income_controlled)
n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)

@@

Selection effects are also different between men and women (Appendix
Figure \@ref(fig:plot-sexes)). Differences are particularly large for
educational attainment, height and MDD. Several PGS for mental illness
and personality traits are more selected for (or less against) among
women, including major depressive disorder (MDD), schizophrenia and
neuroticism, while extraversion is more selected for among men. Waist
circumference and waist-hip ratio are less selected for among women, and
educational attainment is more selected against, though the difference
is only significant for EA3. One possible reason for these sex differences is
that polygenic scores may affect fertility via mate search.

@@

We next focus on variables related to mate search and household
formation. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 partners. Splitting respondents by whether
they were living with a spouse or partner at the time of interview,
effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
respondents by age at first live birth (AFLB).[^1] There is evidence for
genetic effects on AFLB [@Barban_2016], and there is a close link
between this variable and number of children born. Figure
\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
each tercile of AFLB. Several effects are strikingly different across
terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but
is not significantly selected among the youngest third. Similarly,
several PGS of body measurements are selected against only among older
mothers. The correlation between effect sizes for the youngest and
oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
two effects: changing the number of children within each category, and changing
the number of children by shifting respondents between categories.

@@

To investigate this further, we regress number of children on PGS
*controlling* for AFLB (appendix Figure \@ref(fig:plot-age-flb)). In
`r opp_signed` out of `r length(score_names)` cases, effects
change sign when controls are added. The correlation between effect
sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
Selection effects appear to come through two opposing channels: an
effect on AFLB, and an opposite-signed effect on number of children 
controlling for AFLB.

@@

We also check whether selection effects differ by socio-economic status in the
parents' generation. We have no information about parents' income, so we use the
1971 Townsend deprivation score of respondents' birthplace as a proxy
[@townsend1987deprivation]. Results (Figure \@ref(fig:plot-siblings-townsend))
show the same pattern as for respondents: effect sizes are larger and more
significant in the most deprived areas.

@@

Our empirical results can be explained by the economic
theory of fertility [@becker1960economic; @willis1973new; @becker1976child].
According to this, increases in a person's wage affect their fertility via two
opposing channels. There is an *income effect* by which children become more
affordable, like any other good. There is also a *substitution effect*: since
childrearing has a cost in time, the opportunity cost of childrearing increases
if one's market wage is higher. The income effect would lead higher earners to
have more children. The substitution effect would lead higher earners to have
fewer children. If so, then genetic variants which affect earnings potential in
the labour market may cause opposing effects on fertility. The income effect
will cause natural selection in favour of earnings-increasing variants. The
substitution effect will do the reverse.

This theory predicts that genetic variants will be selected for (or against) in
proportion to their effect on earnings. Figure
\@ref(fig:plot-cors-earnings-educ) plots selection effects on each PGS against
that PGS's correlation with earnings in a respondent's first job[^earnings-est],
and against its correlation with educational attainment, a predictor of lifetime
earnings. The raw relationships (left column) are strongly negative. If we plot
selection effects controlling for age at first live birth (right column), the
effect reverses and becomes positive. Thus, the labour market appears to
play an important role in natural selection. Substitution effects
dominate income effects overall, which fits the empirical association 
between income and lower fertility [@becker1960economic;@jones2006economic].

We also re-estimate equation \@ref(eq:regression) for each
PGS, controlling for earnings in first job and education levels. Effect sizes
are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
earnings and education indeed mediate the effect of PGS on fertility.

@@

dfr <- data.frame(
        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
        Controls     = rep(c("None", 
                        "Age at first live birth"), each = 66)
      )

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None")
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        facet_grid(Correlation ~ Controls, labeller = label_both) + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(
          legend.position = "none",
          panel.spacing = unit(14, "pt")
        )

@@

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and age at smoking initiation, are selected for. However, this theory
is less good at explaining variation in selection across the full range of
scores, including physical measures, e.g. waist-hip ratio and BMI (see also
appendix Figure \@ref(fig:plot-together)). Externalizing behaviour also does not
explain why selection effects change sign among older parents.

We test the alternative theory directly by re-estimating equation
\@ref(eq:regression) controlling for a measure of risk attitude (Biobank field
2040). While risk attitude is always a highly significant predictor of number of
children, it has little impact on the effect size or significance of the PGS.
The median ratio of PGS effect sizes between regressions with and without
controls is `r median(ratio)`; all PGS which are significant at $p \le 0.05/33$
in uncontrolled regressions remain so when controlling for risk attitude. Thus,
although risk attitude does predict number of children, it appears not to
mediate selection effects. Overall, we believe that the economic theory is the
leading explanation.

@@

PGS which correlate with high (low) earnings and more (less) education are being
selected against (selected for). In addition, many of the phenotypes under
positive selection are linked to poor health, or are what many people would see
as undesirable to have. For example, most people would probably prefer to have
high educational attainment, a low risk of major depressive disorder, and a low
risk of coronary artery disease, but natural selection is pushing against
genetics linked to these traits. Potentially, this could increase the health
burden on modern populations, but that depends on effect sizes. Our results
suggest that naïve estimates can be affected by sample ascertainment bias. This
problem may be less serious in surveys which aim to be representative (as the UK
Biobank does not). Nevertheless, not all respondents consent to the collection
of genetic data: for instance, completion rates for genotype data in the US
Health and Retirement Study were around 80-85% [@hrs2020]. Researchers should be
aware of the risks of ascertainment when studying modern natural selection.

In addition, we do not know whether estimated effect sizes of natural
selection will increase or decrease as more accurate polygenic scores
are produced. We are also unsure whether genetic variants underlying other
phenotypes will show a similar pattern of natural selection as these ones. 
In short, it is probably too early to tell whether modern natural
selection has a substantively important effect on populations' mean 
polygenic scores.

Because selection effects are concentrated in poorer populations, they also increase
inequality with respect to polygenic scores. For example, Figure 
\@ref(fig:plot-mean-EA3-by-income) graphs children's mean polygenic scores for 
educational attainment (EA3), in households of different income groups. The blue
bars show the actual means. The grey bars show the hypothetical means if all 
households had equal numbers of children. Natural selection against EA3, which
lowers the mean, is stronger at the bottom of the income distribution, and this
increases the differences between groups.

@@

>>>>

---
title: "Labor Market Income and Substitution Effects Explain Modern Natural Selection"
author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against). Selection effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Times
---

@@

* Limit max no. of children in regressions (use an "in_children" bool)
  - not necessary. Very few have abnormally many.
* Consider doing separate regressions for "any children" vs. "n children | 1 or more"
  - heckit too slow and gives crazy answers, just use plm/ols
  - some variables do look different
* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1.
* Reweight other regressions (e.g. the omnibus regression in the appendix)?

* If you take "year of last job started" and see if it is after/before
  the first child, you can figure out whether mothers keep working.
  - better still, look at all jobs started and see if any are within e.g. 5 years of
  first child's birth
  - this could then be another way to get at income vs substitution effects. You'd
  predict negative selection effects among mothers who worked. Of course the
  choice to work is endogenous.
* double check which scores are reverse coded!

@@

As yet we know little about the mechanisms behind these effects. This study uses
data from UK Biobank to learn more. We test for natural selection on 33
different polygenic scores by estimating their correlation with fertility. We
extend the analysis over two generations, using data on respondents' number of
siblings as well as their number of children. This is interesting because
consistent natural selection over multiple generations could lead to substantive
effects in the long run. Most importantly, we examine reproductive rates in
different subgroups of the population, in order to uncover patterns that can
help illuminate the mechanisms behind modern natural selection.

We find selection effects on many polygenic scores. Effects are consistent
across generations. The strength of natural selection on a polygenic score is
associated with that score's correlation with education and earnings: scores
that predict lower education and earnings are being selected for. Also, across
the board, polygenic scores have stronger relationships with fertility among
specific subgroups. Selection effects are stronger among people with lower
income and less education, younger parents, people not living with a partner,
and people with many lifetime sexual partners. Outside these groups, effects are
weaker and often statistically insignificant. In fact, in some subgroups, the
direction of selection is actually reversed: polygenic scores predicting
higher education and earnings are associated with *higher* fertility.

@@

We use data from UK Biobank. Our sample consists of white British (as defined
by genetic estimated ancestry). For regressions on number of children we 
limit the sample to people over 45, since most fertility is completed by this age. We compute 33 polygenic scores, normalized
to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores in the sample by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order
of 5% of a standard deviation.

@@

These changes could reflect natural selection within the UK
population, but also ascertainment bias within the sample. Respondents are
higher income and better educated than the UK population, and they may
also differ on other unobserved characteristics [@10.1093/aje/kwx246].
Since richer and educated people also live longer, this bias might also
increase with age.

To test directly for natural selection, we regress respondents'polygenic scores
on their number of children ($y_i$):

@@

The "selection effect", $\beta$, measures the strength of natural
selection within the sample. In fact, since polygenic scores are
normalized, $\beta$ is the expected polygenic score among
children of the sample [@beauchamp2016genetic].^[normalization] To learn more about the
underlying mechanisms, we split the sample, starting with basic
demographic variables including education, income and sex. These are all
potential sources of ascertainment bias: as well as the ascertainment
for income and education, mentioned above, the sample sex ratio skews
`r female_ratio * 100`% female.

^[normalization]: The selection effect $\beta$ equals $Cov(Y, PGS)/Var(PGS)$
where $Y$ is the number of children. Since PGS are normalized to variance 1,
this reduces to to $Cov(Y, PGS) \equiv E(Y PGS) - E(Y) E(PGS)$, which in turn reduces to $E(Y PGS)$ as $E(PGS) = 0$. This is the polygenic score weighted
by the number of children, which is the average polygenic score in the 
next generation.

@@

```{r calc-income-edu-controlled}

drake::loadd(res_income_controlled, res_edu_controlled)
n_regs_ic <- nrow(res_income_controlled)
n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)

@@

Selection effects are also different between men and women (Appendix
Figure \@ref(fig:plot-sexes)). Differences are particularly large for
educational attainment, height and MDD. Several PGS for mental illness
and personality traits are more selected for (or less against) among
women, including major depressive disorder (MDD), schizophrenia and
neuroticism, while extraversion is more selected for among men. PGS for waist
circumference and waist-hip ratio are less selected for among women, and PGS for
educational attainment are more selected against, though the difference
is only significant for EA3. One possible reason for these sex differences is
that polygenic scores may affect fertility via success in marriage markets, and
men and women may value different characteristics in these.

@@

We next focus on variables related to household type and reproductive
strategy. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
respondents by age at first live birth (AFLB).[^1] There is evidence for
genetic effects on AFLB [@Barban_2016], and there is a close link
between this variable and number of children born. Figure
\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
each tercile of AFLB. Several effects are strikingly different across
terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but
is not significantly selected among the youngest third. Similarly,
several PGS of body measurements are selected against only among older
mothers. The correlation between effect sizes for the youngest and
oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
two effects: changing the number of children within each category, and changing
the number of children by shifting respondents between categories.

@@

To investigate this further, we regress number of children on PGS
*controlling* for AFLB (appendix Figure \@ref(fig:plot-age-flb)). In
`r opp_signed` out of `r length(score_names)` cases, effects
change sign when controls are added. The correlation between effect
sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
Thus, selection effects seem to come through two opposing channels: an
effect on AFLB, and an opposite-signed effect on number of children 
controlling for AFLB.

@@

We also check whether selection effects differ by socio-economic status in the
parents' generation. We have no information about parents' income, so we use the
1971 Townsend deprivation score of respondents' birthplace as a proxy
[@townsend1987deprivation]. Results (Figure \@ref(fig:plot-siblings-townsend))
show the same pattern as for respondents: effect sizes are larger and more
often significant in the most deprived areas.

@@

# Economic theory and natural selection

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. This can potentially be explained by the
economic theory of fertility [@becker1960economic; @willis1973new;
@becker1976child]. According to this, increases in a person's wage affect their
fertility via two opposing channels. There is an *income effect* by which
children become more affordable, like any other good. There is also a
*substitution effect*: since childrearing has a cost in time, the opportunity
cost of childrearing increases if one's market wage is higher. The income effect
would lead higher earners to have more children. The substitution effect would
lead higher earners to have fewer children. If so, then genetic variants which
affect earnings potential in the labour market may cause opposing effects on
fertility. The income effect will cause natural selection in favour of
earnings-increasing variants. The substitution effect will do the reverse.

@@

dfr <- data.frame(
        PGS          = rep(score_names, 4),
        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
        Controls     = rep(c("None", 
                        "Age at first live birth"), each = 66)
      )

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None"),
        PGS      = ifelse(abs(effect_size) > 0.035, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.012) +
        facet_grid(Correlation ~ Controls, labeller = label_both) + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        scale_colour_brewer(type = "qual", palette = 2) +
        coord_cartesian(clip = "off") +
        theme(
          legend.position = "none",
          panel.spacing = unit(14, "pt")
        )

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each PGS against that PGS's correlation with earnings in a respondent's first
job[^earnings-est], and against its correlation with educational attainment, a
predictor of lifetime earnings. The raw relationships (left column) are strongly
negative. If we plot selection effects controlling for age at first live birth
(right column), the effect reverses and becomes positive. Thus, the labour
market appears to play an important role in natural selection. Substitution
effects dominate income effects overall, which fits the known association
between income and lower fertility [@becker1960economic;@jones2006economic].

Second, we re-estimate equation \@ref(eq:regression) for each
PGS, controlling for earnings in first job and education levels. Effect sizes
are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
earnings and education indeed mediate the effect of PGS on fertility.

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and age at smoking initiation, are selected for. However, this theory
is less good at explaining variation in selection across the full range of
scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
behaviour also does not explain why selection effects should work in the
opposite direction among older parents, whereas the economic theory can explain
this via income effects, as described above. We test the alternative theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (Biobank field 2040). While risk attitude is always a
highly significant predictor of number of children, it has little impact on the
effect size or significance of the PGS. The median ratio of PGS effect sizes
between regressions with and without controls is `r median(ratio)`; all PGS
which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
when controlling for risk attitude. Thus, although risk attitude does predict
number of children, it appears not to mediate selection effects. Overall, we
believe that the economic theory is the leading explanation.

@@

PGS which correlate with high (low) earnings and more (less) education are being
selected against (selected for). In addition, many of the phenotypes under
positive selection are linked to poor health, or are what many people would see
as undesirable to have. For example, most people would probably prefer to have
high educational attainment, a low risk of major depressive disorder, and a low
risk of coronary artery disease, but natural selection is pushing against
genetics linked to these traits. Potentially, this could increase the health
burden on modern populations, but that depends on effect sizes. Our results
suggest that naïve estimates can be affected by sample ascertainment bias. This
problem may be less serious in surveys which aim to be representative (as the UK
Biobank does not). However, there is still scope for bias, since not all
respondents consent to the collection of genetic data. For instance, completion
rates for genotype data in the US Health and Retirement Study were around 80-85%
[@hrs2020]. Researchers should be aware of the risks of ascertainment when
studying modern natural selection.

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced. And we are unsure
whether genetic variants underlying other phenotypes will show a similar pattern
of natural selection to those studied here. In short, it is probably too early
to tell whether modern natural selection has a substantively important effect on
population genetics. Nevertheless, we note that selection effects on our measured
polygenic scores are still relatively small, even after reweighting to account for ascertainment bias.

Because selection effects are concentrated in poorer groups, they also increase
inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3), for children in households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against EA3, which lowers the
mean, is stronger at the bottom of the income distribution, and this increases
the differences between groups.

@@

## Effects on inequality

Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean polygenic
scores between the highest and lowest income groups. Column "Actual" shows parents' scores weighted by Age/Qualification times number of children. 
Column "Without selection" shows scores weighted by Age/Qualification only, i.e. if all couples had the same number of children.

```{r tbl-inequality-effects}

drake::loadd(age_qual_weights)

ineq <- famhist %>% 
          left_join(age_qual_weights, by = "f.eid") %>% 
          filter(! is.na(n_children), ! is.na(income_cat), ! is.na(weights)) %>% 
          group_by(income_cat) %>% 
          summarize(across(all_of(score_names), 
                    list(
                      unw = ~ weighted.mean(.x, weights, na.rm = TRUE),
                      wtd = ~ weighted.mean(.x, weights * n_children, 
                                              na.rm = TRUE)
                    )
          )) %>% 
          tidyr::pivot_longer(-income_cat, 
                                names_to = c("score", ".value"), 
                                names_pattern = "(.*)_(.+?)$"
                              )

ineq_comp <-  left_join(
                ineq %>% filter(income_cat == 1),
                ineq %>% filter(income_cat == 5),
                by = "score"
              )

ineq_comp %<>% mutate(
                 diff_children = wtd.y - wtd.x,
                 diff_hypoth   = unw.y - unw.x
               )

ineq_comp %>% 
      select(
        PGS                 = score, 
        `With selection`    = diff_children, 
        `Without selection` = diff_hypoth
      ) %>% 
      mutate(
        PGS = pretty_names(PGS)
      ) %>% 
      arrange(desc(Actual)) %>% 
      as_hux() %>% 
      set_align(everywhere, -1, "right") %>% 
      set_caption("Differences in polygenic scores between highest and lowest income group") %>% 
      theme_compact()
```
